---
title: "Data 200: Lecture 8"
subtitle: "Introduction to Machine Learning: Simple linear regression"
author: "Medina - CSU Channel Islands"
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
execute:
  echo: true 
---

```{r}
#| echo: false
library(janitor)
library(broom)
library(gridExtra)
library(tidyverse)

options(scipen = 999)

theme_set(theme_bw(base_size = 22))
```

# Machine learning

**Machine learning** = Methods that use algorithms to learn patterns from data and make predictions

- Simple linear regression is one example of a machine learning method

## Data `babies` in `openintro` package

```{r}
library(openintro)

data("babies")

babies <- babies |> 
  filter(!is.na(bwt) & !is.na(gestation)) 

glimpse(babies)
```

##  Baby Weights


```{r}
#| output-location: column

babies |> 
  ggplot(
    aes(x = gestation, y = bwt)
  ) +
  geom_point() +
  labs(
    x = "Gestation (days)",
    y = "Birthweight (oz)"
  )

```


##  Baby Weights


```{r }
#| output-location: column

babies |> 
  ggplot(
    aes(x = gestation, y = bwt)
  ) +
  geom_point() +
  labs(
    x = "Gestation (days)",
    y = "Birthweight (oz)"
  ) +
  geom_smooth(
    method = "lm", 
    se = FALSE
  ) 
```

`lm` stands for linear model  
`se` stands for standard error

##  Baby Weights


```{r }
#| echo: false

babies |> 
  ggplot(
    aes(x = gestation, y = bwt)
  ) +
  geom_point() +
  labs(
    x = "Gestation (days)",
    y = "Birthweight (oz)"
  ) +
  geom_smooth(
    method = "lm", 
    se = FALSE
  ) 
```

The line is our "model". If we can describe a phenomena (e.g., relationship between birth weight and gestation) with a model, we can use that model 

- to describe relationships
- to predict future values

## Notation

<center>


| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Birth weight | y | Response | Numeric |
| Gestation | x | Explanatory | Numeric |

</center>

. . .

How do we get and use the line?


## Linear Equations Review


:::{.pull-left}

Recall from your previous math classes

$y = mx + b$

where $m$ is the slope and $b$ is the y-intercept

e.g. $y = 2x -1$
:::

. . .



:::{.pull-right}

```{r}
#| echo: false
x <- c(0, 1, 2, 3, 4, 5)
y <- c(-1, 1, 3, 5, 7, 9)

as.data.frame(x = x, y = y) |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  scale_y_continuous(breaks = c(-1, 1, 3, 5, 7, 9)) +
  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5)) +
  geom_smooth(method = "lm", se = FALSE)

```

:::


## Linear model

:::{.pull-left}

**Math** class

$y = b + mx$

- $b$ is y-intercept  
- $m$ is slope  
:::


:::{.pull-left}

**Data science** class

$y_i = b_0 +b_1x_i + \epsilon_i$

- $b_0$ is **y-intercept**
- $b_1$ is **slope**  
- $\epsilon_i$ is **leftover variation / unexplained randomness**  
- $i = 1, 2, ...n$ identifier for each point
:::

## Linear model

:::{.pull-left}

**Data science** class

$y_i = b_0 +b_1x_i + \epsilon_i$

- $b_0$ is **y-intercept**
- $b_1$ is **slope**  
- $\epsilon_i$ is **leftover variation / unexplained randomness**  
- $i = 1, 2, ...n$ identifier for each point 
:::


:::{.pull-left}

Note $y_i = b_0 +b_1x_i + \epsilon_i$ is the **theoretical model**. 

- In reality we don't have data for everything / everyone we want to learn about, so we use a sample to get estimates for $b_0$ and $b_1$.

:::


## How to fit a linear model with R

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)
```

- the `lm()` function fits a linear model, often called **regression**.
- `lm` stands for linear model.
- Note that the variables are entered in y ~ x order, where we think x eXplains y



## Model output

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)
broom::tidy(linear_baby_model)
```

. . .

**Estimated model** (general form):

$\hat {y}_i = \hat{b}_0 + \hat{b}_1 x_i$ <- *The hat means it is an estimate!*

. . .

**Estimated model** (context specific):

$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464\text{ gestation}_i$

# We can use our model for prediction!

## Predicted birth weight for a baby with 300 days of gestation

\begin{equation*}
  \begin{split}
    \widehat{\text{Birth Weight}_i} &= -10.1 + 0.464\text{ gestation}_i\\ 
    &= -10.1 + 0.464 (300)\\ 
    &=129.1
  \end{split}
\end{equation*}

For a baby with 300 days of gestation the predicted birth weight is `r -10.1 + 0.464*300` ounces.

## Predicted birth weight for a baby with 300 days of gestation


```{r}
# Define x value for data to predict with
prediction_data <- data.frame(gestation = 300)

predict(linear_baby_model, newdata = prediction_data)
```


```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6
#| fig-align: center

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(
    data = mutate(prediction_data, bwt = predict(linear_baby_model, newdata = prediction_data)), 
    color = "red",
    size = 2
  )
```


## Interpretation of estimates: general

- $\hat{b}_0$ is the predicted/expected $Y$ value when $X = 0$.

- $\hat{b}_1$ is the difference in predicted/expected $Y$ value associated with a 1 unit increase in $X$

## Interpretation of estimates: in context


:::{.pull-left}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 
```

$\hat{b}_1 = 0.464$ which means for one unit (day) increase in gestation period the expected increase in birth weight is 0.464 ounces.

:::

. . .

:::{.pull-right}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0, 360) +
  ylim(-10, 180) +
  geom_abline(slope = 0.459, intercept = -8.76)
```

$\hat{b}_0 = -10.1$ which means for a gestation period of 0 days the expected birth weight is -10.1 ounces!!!!!!!!
(does NOT make sense)
:::



## Extrapolation

- There is no such thing as 0 days of gestation.

. . .

- Birth weight cannot possibly be -10.1 ounces.

. . .

- Extrapolation happens when we use a model outside the range of the x-values that are observed. After all, we cannot really know how the model behaves (e.g. may be non-linear) outside of the scope of what we have observed. 

# Model Error

## Baby number 148

Let's consider a baby in our data and see how far/close the true birth weight is from the predicted birth weight.

. . .

:::{.pull-left}

```{r}
babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)
```

:::

:::{.pull-right}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6

baby_148 <- babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red")
```


:::



## Baby #148

**Observed**:

$$\text{Birth Weight}_{148} = 160$$

**Predicted**:

\begin{equation*}
  \begin{split}
    \widehat{\text{Birth Weight}_i} &= -10.1 + 0.464\text{ gestation}_i\\ 
    &= -10.1 + 0.464 (300)\\ 
    &=129.1
  \end{split}
\end{equation*}



## Residual for `i = 148`

:::{.pull-left}

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red", size = 2) +
  geom_segment(
    x = 300, xend = 300, y = 128.94, yend = 160, 
    color = "red",
    size = 2,
    linetype = "dotted"
  )
```



:::

:::{.pull-right}

$\text{Birth Weight}_{148} = 160$

$\widehat{\text{Birth Weight}}_{148} = 129.1$

:::

**Residual:**

$e_{i} = y_{i} - \hat y_{i}$

$e_{148} =$ 160 - 129.1 =  30.9


## Least Squares Regression 

There are different ways we could have fit a line to our data, we choose the one ($\hat{b}_0$ and $\hat{b}_1$) that minimizes the **sum of squared residuals** ($e_i = y_{i} - \hat y_{i}$):

$$e_1^2 + e_2^2 + ... + e_n^2$$

. . .

*Why didn't we choose to minimize*

$$e_1 + e_2 + ... + e_n?$$

## Prediction Error

**Root Mean Squared Error** (RMSE) can be thought of as "average error" of our model

$$RMSE = \sqrt{(e_1^2 + e_2^2 + ... + e_n^2) / n},$$
with $e_i = y_{i} - \hat y_{i}$.

. . .

- Minimum value for RMSE is 0
- RMSE is in the same units as your $Y$ variable


## Computing residuals

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)

babies |> 
  modelr::add_predictions(linear_baby_model) |> 
  modelr::add_residuals(linear_baby_model) |> 
  select(gestation, bwt, pred, resid)
```

## Computing RMSE

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)

babies <- babies |> 
  modelr::add_predictions(linear_baby_model) |> 
  modelr::add_residuals(linear_baby_model)

sqrt(mean((babies$bwt - babies$pred)^2))
```

Our linear model's RMSE is `r round(sqrt(mean((babies$bwt - babies$pred)^2)), 2)` oz.

. . .

*Maybe weight of the mother would better predict baby birth weight... We could fit both linear models and compare prediction error!*
