---
title: "DATA 200: Lab 6b"
format: 
  html:
    embed-resources: true
execute:
  message: false
  warning: false
  eval: false
  echo: true
---

# National park visitation data

For today's lab we will combine many of the coding skill we have used throughout this course to explore US National Park Service [visitor data](https://catalog.data.gov/dataset/nps-visitor-use-statistics-data-package-2025).

You will first need to prepare the data, then we can summarize and plot it.
This is meant to provide you with practice on conducting exploratory data analyses.
**Do not use any resources for this lab aside from your notes, our slides, or our textbook.**

```{r}
library(tidyverse)
library(janitor)
```

```{r}
park_visits_og <- read_csv(("https://csuci-data-200-f26.github.io/lecture-materials/data/Main_State_Data.csv"))
```

```{r}
park_visits_og |> 
  clean_names() |> 
  filter(state_id == "CA") |> 
  select(unit_code) |> 
  unique()
```


```{r}
park_unit_codes <- read_csv(("https://csuci-data-200-f26.github.io/lecture-materials/data/california_nps_unit_codes.csv"))
```

## Preparing our data

*30 minutes you try then 10 minutes we try.*

0. Make a quarto doc in RStudio called "lab-06b.qmd" and save it into your week 6 repo.
1. Read in the park visit data.
2. Change the column names to comply with tidyverse style guidelines.
3. Read in the unit code data.
4. Join the two datasets, keeping only rows with a unit code available in both datasets, and save the joined data into an object called `park_visits`.
5. Filter to recreation visits only.
6. Convert year and month to be factors.
7. Create a new variable called `date`. We have `year` and `month` available to us in separate columns, so we can't use the `myd()` functions like we've learned. Instead we can use the `make_date()` function, providing it the `year` and `month` columns, and for day we can say `1`.


## Analyzing our data

*15 minutes you try then 10 minutes we try.*

### Summarizing our data

1. Calculate total number of recreation visits for each park and year. Note that the `value` column has counts, so within our `summarize()` we will want `sum(value)`. Which California park has the most and fewest the most visits?

### Visualizing our data

1. Choose 5 California parks you are interested in investigating.
2. Filter your data to only contain those 5 parks
3. Make a line plot of number of recreation visits by date with each park mapped to color. 
4. Improve your plot by employing accessibility considerations and provide informative labels.


## A problem-solving workflow

When you encounter a data problem:

1. Identify what is preventing you from obtaining the desired result.
2. Describe the transformation you need in ordinary language. Try writing it out in words.
3. Search for a function or example that performs that transformation.
5. Try out a solution and inspect the result.
5. Decide whether it actually solved the problem. If not, reiterate.

