---
title: "Lab 6b"
format: html
---

## A problem-solving workflow

When you encounter a data problem:

1. Identify what is preventing you from obtaining the desired result.
2. Describe the transformation you need in ordinary language. Try writing it out in words.
3. Search for a function or example that performs that transformation.
5. Try out a solution and inspect the result.
5. Decide whether it actually solved the problem. If not, reiterate.

## National park visitation data

```{r}
unit_codes <- daata.frame(Statistic,StatisticDescription
TRV,Recreation Visits
TNRV,Non-Recreation Visits
TV,Total Visits
TRVH,Recreation Visitor Hours
TNRVH,Non-Recreation Visitor Hours
TH,Visitor Hours (Recreation plus Non-Recreation)
TT,NPS Operated Campground Tent Overnight Stays
TRVS,NPS Operated Campground RV Overnight Stays
TTRV,NPS Operated Campground Overnight Stays (Tent plus RV)
CL,Concessionaire Operated Lodging Overnight Stays
CCG,Concessionaire Operated Campground Overnight Stays
BC,Backcountry Overnight Stays
NROS,Non-Recreation Overnight Stays
MISC,Miscellaneous Overnight Stays
TOS,Total Overnight Stays
```


### Statistics

| Statistic | Statistic Description |
|:---------|:-------------------------|
| TRV | Recreation Visits |
| TNRV | Non-Recreation Visits |
| TV | Total Visits |
| TRVH | Recreation Visitor Hours |
| TNRVH | Non-Recreation Visitor Hours |
| TH | Visitor Hours (Recreation plus Non-Recreation) |
| TT | NPS Operated Campground Tent Overnight Stays |
| TRVS | NPS Operated Campground RV Overnight Stays |
| TTRV | NPS Operated Campground Overnight Stays (Tent plus RV) |
| CL | Concessionaire Operated Lodging Overnight Stays |
| CCG | Concessionaire Operated Campground Overnight Stays |
| BC | Backcountry Overnight Stays |
| NROS | Non-Recreation Overnight Stays |
| MISC | Miscellaneous Overnight Stays |
| TOS | Total Overnight Stays | 

```{r}
library(tidyverse)
library(janitor)
```

```{r}
park_visits <- read_csv(here::here("data/Main_State_Data.csv")) |> 
  clean_names()
```

```{r}
glimpse(park_visits)
```

```{r}
cal_parks <- park_visits |> 
  filter(unit_code %in% c("CHIS", "JOTR", "REDW", "YOSE"))

glimpse(cal_parks)
```

Cleaning our data:

1. Filter to 5 parks of interest.
2. Filter to recreation visits only.
3. Need to convert year and month to be factors.
4. Create a new variable called `date`. We have `year` and `month` available to us in separate columns, so we can't use the `myd()` functions like we've learned. Instead we can use the `make_date()` function, providing it the `year` and `month` columns, and for day we can say `1`.

Analyzing our data:

1. Calculate total number of recreation visits for each park and year. Note that the `value` column has counts, so within our `summarize()` we will want `sum(value)`.
2. Make a scatter or line plot of number of recreation visits by date for each park.
3. Make a column plot (like a bar plot, but where the data is already aggregated `geom_col()`) of number of reacte

```{r}
cal_parks |> 
  filter(statistic == "TRV") |> 
  group_by(unit_code, year) |> 
  summarize(
    sum(value)
  )
```



```{r}
#| eval: false
cal_parks |> 
  filter(statistic == "TRV") |> 
  mutate(month = as.factor(month), year = as.factor(year)) |> 
  ggplot(aes(x = year, y = value, fill = month)) +
  geom_col() +
  scale_fill_viridis_d() +
  theme_minimal()
```

```{r}
cal_parks |> 
  filter(statistic == "TRV") |> 
  mutate(month = as.factor(month)) |> 
  mutate(date = make_date(year, month, 1)) |> 
  ggplot(aes(x = date, y = value, color = unit_code)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_minimal()
```

