---
title: ""
format: 
  pdf:
    echo: true
    message: false
    warning: false
    geometry:
      - bottom=1in
header-includes:
  - \usepackage{lastpage}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{Data 200 Fall 2026 -- \textbf{Lab 6a}}
  - \fancyhead[R]{Factors, Strings, and Dates}
  - \fancyfoot[C]{\thepage\ of \pageref{LastPage}}
  - \renewcommand{\headrulewidth}{0.4pt}
---

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

**Team members**:


```{r}
#| echo: false
library(tidyverse)
```

Consider the following artificial dataset:

```{r}
#| echo: false
customers <- data.frame(
  first_name = c("Lee", "Aaliyah", "EMILY", "Sam", "omar"),
  satisfaction = c("high", "med", "high", "low", "med"),
  signup_date = c("01/03/2024", "05/30/2024", "02/15/2024", "03/10/2024", "01/22/2024")
) 
```

```{r}
#| echo: false
knitr::kable(customers)
```

**Question 1**\newline
For each column, write the variable type and the most appropriate R storage type.

- `first_name`:

\vspace{0.75cm}

- `satisfaction`:

\vspace{0.75cm}

- `signup_date`:

\vspace{0.75cm}

**Question 2**\newline
You should have identified that one of the columns would be best stored as a `factor`. Why would it be better stored as a `factor` rather than a `character`?

\vspace{2.25cm}

**Question 3**\newline
You should have identified that one of the columns would be best stored as a `date`. Why would it be better stored as a `date` rather than a `character`?

\vspace{2.25cm}

**Question 4**\newline
You should have identified that one of the columns would be best stored as a `character`. Why would it be better stored as a `character` rather than a `factor`?

\vspace{2.25cm}

**Question 5**\newline
Complete the following R code to change each of the columns to the ideal R storage type.
```{r}
#| eval: false
customers <- 
  customers |> 
  mutate(
    first_name = 
      
    satisfaction = 
      
    signup_date = 
  )
```

**Question 6**\newline
By default, variables stored as factors will have the levels stored in alphabetical order. Complete the R code below to convert the `satisfaction` variable to a factor with a more meaningful ordering and labels appropriate for a plot.
```{r}
#| eval: false
customers <- 
  customers |> 
  mutate(________________ = factor(
    
    levels = c(__________, __________, __________),
    
    labels = c(_______________, _______________, _______________)
  ))
```

**Question 7**\newline
Describe what the following code seems to be doing.

```{r}
#| echo: false
customers <- customers |> 
  mutate(signup_date = mdy(signup_date))
```

```{r}
customers |> 
  mutate(membership_length = ymd("2025-02-23") - signup_date) |> 
  arrange(membership_length)
```

\newpage

**Question 8**\newline
Make a new column called `membership_month` which stores the month that the customer signed up. (Hint: use the `month()` function.)

\vspace{4cm}

**Question 9**\newline
Which of the columns from the original dataset could be described as a "string"?

\vspace{1cm}

**Question 9**\newline
Describe what the following code seems to be doing.

```{r}
customers |>
  filter(first_name != str_to_title(first_name)) |>
  pull(first_name)
```

\vspace{3cm}

**Question 10**\newline
Oh no! It looks like names were not recorded with consistent capitalization. We want them to be recorded title case, meaning the first letter should be capital, followed by lower case letters. Write R code to store the names in title case. (Hint: use the `str_to_title()` function). Make sure to save the change back into the data frame.

