---
title: ""
format: 
  pdf:
    echo: true
    message: false
    warning: false
    geometry:
      - bottom=1in
header-includes:
  - \usepackage{lastpage}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{Data 200 Fall 2026 -- \textbf{Lab 6a}}
  - \fancyhead[R]{Factors, Strings, and Dates}
  - \fancyfoot[C]{\thepage\ of \pageref{LastPage}}
  - \renewcommand{\headrulewidth}{0.4pt}
---

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

**Team members**:


```{r}
#| echo: false
library(tidyverse)
```

Consider the following artificial dataset for gym memberships:

```{r}
#| echo: false
customers <- data.frame(
  id = c(101, 102, 103, 104, 105),
  first_name = c("Lee", "Aaliyah", "EMILY", "Sam", "omar"),
  visit_frequency = c("high", "med", "high", "low", "med"),
  signup_date = c("01/03/2026", "05/30/2026", "02/15/2026", "08/10/2026", "09/22/2026")
) 
```

```{r}
#| echo: false
knitr::kable(customers)
```

**Question 1**\newline
For each column, write the variable type and the most appropriate R storage type.

- `first_name`:

\vspace{0.75cm}

- `visit_frequency`:

\vspace{0.75cm}

- `signup_date`:

\vspace{0.75cm}

**Question 2**\newline
You should have identified that one of the columns would be best stored as a `factor`. Why would it be better stored as a `factor` rather than a `character`?

\vspace{2.25cm}

**Question 3**\newline
You should have identified that one of the columns would be best stored as a `date`. Why would it be better stored as a `date` rather than a `character`?

\vspace{2.25cm}

**Question 4**\newline
You should have identified that one of the columns would be best stored as a `character`. Why would it be better stored as a `character` rather than a `factor`?

\vspace{2.25cm}

**Question 5**\newline
Which of the columns from the original dataset could be described as a "string"?

\vspace{1cm}

**Question 6**\newline
Complete the following R code to change each of the columns to the ideal R storage type.
```{r}
#| eval: false
customers <- 
  customers |> 
  mutate(
    first_name = 
      
    visit_frequency = 
      
    signup_date = 
  )
```

**Question 7**\newline
Oh no! It looks like names were not recorded with consistent capitalization. We want them to be recorded title case, meaning the first letter should be capital, followed by lower case letters. Write R code to store the names in title case. (Hint: use the `str_to_title()` function). Make sure to save the change back into the data frame.

\vspace*{1.5in}

**Question 8**\newline
By default, variables stored as factors will have the levels stored in alphabetical order. Complete the R code below to convert the `visit_frequency` variable to a factor with levels ordered by "low", "med", "high".
```{r}
#| eval: false
customers <- 
  customers |> 
  mutate(
    ____________________ = fct_relevel(
      ____________________, 
      ____________________, 
      ____________________, 
      ____________________
    )
  )
```

**Question 9**\newline
Actually, I decided I want the reverse order for `visit_frequency`, from high to low. Complete the R code below.
```{r}
#| eval: false
customers <- 
  customers |> 
  mutate(____________________ = ____________________(____________________))
```

**Question 10**\newline
Describe what the following code seems to be doing.

```{r}
#| echo: false
customers <- customers |> 
  mutate(signup_date = mdy(signup_date))
```

```{r}
customers |> 
  mutate(membership_length = ymd("2026-09-28") - signup_date) |> 
  arrange(membership_length)
```

\vspace{1in}

**Question 11**\newline
Make a new column called `membership_month` which stores the month that the customer signed up. (Hint: use the `month()` function.)

\newpage

**Question 12**\newline

\vspace*{-0.5cm}

```{r}
#| echo: false
prices <- tibble(
  id = c(100, 101, 102, 103),
  monthly_price = c(15, 15, 25, 20)
)

knitr::kable(
  list(
    customers = customers,
    prices = prices
  ),
  booktabs = TRUE
)
```

Suppose I now want to join my `customers` dataset with the new `prices` dataset.
Write out what the output would be for each type of join below.

**Question 12a**\newline

\vspace*{-1cm}

```{r}
#| eval: false
full_join(customers, prices)
```

\vspace*{2in}

**Question 12b**\newline

\vspace*{-1cm}

```{r}
#| eval: false
inner_join(customers, prices)
```

\vspace*{1.25in}

**Question 12c**\newline

\vspace*{-1cm}

```{r}
#| eval: false
left_join(customers, prices)
```

