---
title: ""
format: 
  pdf:
    echo: true
    message: false
    warning: false
header-includes:
  - \usepackage{lastpage}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{Data 200 Fall 2026 -- \textbf{Lab 3a}}
  - \fancyhead[R]{Hello Data Science Chapter 3}
  - \fancyfoot[C]{\thepage\ of \pageref{LastPage}}
  - \renewcommand{\headrulewidth}{0.4pt}
---

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

Team members:

\vspace*{1cm}

In this lab we will use a dataset of movies. For each movie we have information on their genre, critic score (from 0-100), rating (G, PG, PG-13, or R), and box office earnings (in millions of dollars). 

```{r}
#| echo: false
library(openintro)
library(tidyverse)

data("movies")
```

## Question 1

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 6
#| fig-height: 2.5

ggplot(movies, aes(x = score, y = rating)) +
  geom_boxplot()
```

List 3 factual observations about the plot shown above.

1.

\vspace{1cm}

2.

\vspace{1cm}

3.

\vspace{1cm}


## Question 2

Which ratings categories appear to be similar in critic score? Which appear to be the least similar? Does this make sense?

\vspace*{2cm}

## Question 3

Identify each plotting step from the code (i.e., circle and number the step):

1. Call `ggplot()` with the data.
2. Specify variables to be mapped to aesthetics (e.g., axes, colors, etc)
3. Add a layer to specify the plot type.

\vspace{0.5cm}

```{r}
#| fig-width: 4
#| fig-height: 3
#| fig-align: center
ggplot(movies, aes(x = box_office)) +
  geom_histogram()
```

\vspace*{-0.5in}

## Question 4

How would you describe the distribution visualized in question 3 (right-skewed, left-skewed, symmetric)?

## Question 5a 

If we doubled the dataset by duplicating every row, how would the plot from question 1 change, if at all?

\vspace{3.5cm}


## Question 5b

If we doubled the dataset by duplicating every row, how would the plot from question 3 change, if at all?

\vspace{3.5cm}



## Question 6

I want to produce a bar plot of movie ratings. Identify which code I should use.

```{r}
#| eval: false

ggplot(movies, x = rating) +
  geom_bar()
```

\vspace*{-0.5cm}

```{r}
#| eval: false

ggplot(aes(movies), x = rating) +
  geom_bar()
```

\vspace*{-0.5cm}

```{r}
#| eval: false

ggplot(movies, aes(x = rating))
  geom_bar()
```

\vspace*{-0.5cm}

```{r}
#| eval: false

ggplot(movies, aes(x = rating)) +
  geom_bar()
```


\newpage

## Question 7

Work with your team to sketch a box plot of box office earnings using information provided by the plot in question 3.

\vspace*{7cm}

## Question 8

I want to investigate which genre (action, comedy, or horror) had more rated R movies and which genre had the fewest PG movies. What type of plot would you recommend? 

Create a sketch of what this plot might look like. (I haven't shown you any data for genre, so this exercise is for you to think about what would be on the x-axis, y-axis, legend, and what the data could look like).

 
