---
title: "Data 200: Lab 1a"
subtitle: "Popular Baby Names"
author: "Your name here"
lang: en
format: 
  html:
    echo: false
    message: false
    warning: false
    toc: true
    toc-depth: 4
    embed-resources: true
---

## Preface 

This file is a **Quarto document** and it allows us to combine text and code.
By clicking **Render**, an html will be produced with all of the output from our code included.
For today we will not worry about how the code, instead we will focus on our investigation.

Start by clicking `Render` and examine the html produced.

In this lab we will explore popular baby names for the state of California.
This data comes from California birth certificates and is available on the [California Open Data Portal](https://data.ca.gov/dataset/most-popular-baby-names).

## Obtain and familiarize yourself with the data

```{r}
library(tidyverse)
```

This data is read in from a file called `california-popular-baby-names.csv` available online.
We will save this data into an **object** called `baby-names`.
```{r}
baby_names <- read_csv("http://csuci-data-200-f26.github.io/lecture-materials/data/california-popular-baby-names.csv")
```

Let's look at what is in our `baby_names` object.
```{r}
baby_names
```

This dataset has 6 columns and 3,254 rows in total, which would be a lot to view at once, so only the first 10 are displayed.


For each year and specified sex at birth this data contains the 25 most popular baby first names and counts.
We have data for each year from 1960 to 2024.

Let's use this data to explore some questions.

### Question 1 

What were the 5 most popular baby names in 2024?

```{r}
year_to_plot <- 2024

baby_names |> 
  filter(Year == year_to_plot) |>
  arrange(desc(Count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = Count, y = reorder(Name, Count))) +
  geom_count() +
  geom_text(aes(label = Count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born",
    y = "Name",
    title = paste0("Top 20 most popular baby names in ", year_to_plot)
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))
```

#### Response

[Replace this text in the brackets with your response to Question 1. To do this you have to edit the Quarto document, like you would with a Word Document, then Render the document to view the updated html file.]

### Question 2

What were the 5 most popular baby names in the year you were born? To investigate this we will change the code, specifically the line that says `year_to_plot <- 2024`. Change the "2024" in the code chunk below in the Quarto document to the year you were born and render the html again. 

```{r}
year_to_plot <- 2024

baby_names |> 
  filter(Year == year_to_plot) |>
  arrange(desc(Count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = Count, y = reorder(Name, Count))) +
  geom_count() +
  geom_text(aes(label = Count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born",
    y = "Name",
    title = paste0("Top 20 most popular baby names in ", year_to_plot)
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))
```


#### Response

[Write your answer to question 2 here.]

### Question 3

Below are all of the top 25 most popular baby names for each year from 1960-2024.
Is your name in the list?

```{r}
sort(unique(baby_names$Name))
```

#### Response

[Write your response to question 3 here.]

### Question 4

What are the top 5 most popular baby names overall from all babies born between 1960-2024?

```{r}
all_names_plot <- baby_names |>
  group_by(Name) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = total_count, y = reorder(Name, total_count))) +
  geom_count() +
  geom_text(aes(label = total_count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born from 1960-2023",
    y = "Name"
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))

all_names_plot
```


#### Response

[Write your response to question 4 here.]

## Correcting truncated names

If you look into the [data dictionary](https://data.ca.gov/dataset/most-popular-baby-names/resource/a99404d8-e3b6-455c-9916-41075f001862) it says:
*Names in records from data years 1960 through 1977 are truncated to 8 characters, and 1978 through 2017 are truncated to 12 characters.* 
*For example, the names CHRISTINA, CHRISTINE, and CHRISTIN would all be included in CHRISTIN*.*

When we look at names across the entire time period we may get different results if we truncated all names to 8 characters.
This truncation is accounted for in the plot below.

### Question 5

Did the top 5 most popular baby names change once we corrected for inconsistent truncation of long names?

```{r}
baby_names <- baby_names |> 
  mutate(name_fixed = case_when(
    str_length(Name) > 8 ~ (str_trunc(Name, width = 9, ellipsis = "*")),
    str_length(Name) <= 8 ~ Name
  )) 

corrected_all_names_plot <-  baby_names |> 
  group_by(name_fixed) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = total_count, y = reorder(name_fixed, total_count))) +
  geom_count() +
  geom_text(aes(label = total_count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born from 1960-2023",
    y = "Corrected names"
  )  +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))

gridExtra::grid.arrange(
  all_names_plot + 
    labs(
      y = "Uncorrect names",
      x = "Number of babies born from 1960-2023"
    ), 
  corrected_all_names_plot, 
  ncol = 2
)
```


#### Response

[Write your response to question 4 here.]


## 5 most popular baby names over the years

```{r}
most_popular_names <- baby_names |> 
  group_by(name_fixed) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:5) |> 
  pull(name_fixed)

baby_names |> 
  filter(name_fixed %in% most_popular_names) |> 
  ggplot(aes(x = Year, y = Count, color = name_fixed, linetype = name_fixed)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Number of babies born",
    color = "Name",
    linetype = "Name"
  ) +
  ylim(0, 9000)
```

### Question 6

Did the most popular name in 1960 stay the most popular? If not, approximately what year did a different name become more popular and what was it?


#### Response

[Write your response to question 6 here.]

### Question 7

Notice that the line for the name "Matthew" does not start until approximately the year 1968. Why do you think that is?

#### Response

[Write your response to question 7 here.]
