---
title: ""
format: 
  pdf:
    echo: true
    message: false
    warning: false
    geometry:
      - bottom=1in
header-includes:
  - \usepackage{lastpage}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{Data 200 Fall 2026 -- \textbf{Lab 5a}}
  - \fancyhead[R]{Hello Data Science Chapter 5}
  - \fancyfoot[C]{\thepage\ of \pageref{LastPage}}
  - \renewcommand{\headrulewidth}{0.4pt}
---

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

**Team members**:


```{r}
#| echo: false
library(tidyverse)
library(hellodatascience)
library(knitr)

data(planets)
```

In this worksheet we will work with the `planets` data frame from the the hellodatascience R package, containing data on planets in our solar system:

```{r}
#| echo: false
kable(planets, format = "latex", booktabs = TRUE)
```

- `name`: Name of planet
- `mass`: Mass in 10^24 kg
- `length_of_day`: Length of day in hours
- `mean_temp`: Whether mean temperature in Celsius is positive or not
- `n_moons`: Number of moons
- `ring_system`: Whether the planet has a set of rings around it
- `surface_pressure`: Surface pressure in bars

**Question 1**\newline
Describe what the `|>` operator does.

\vspace{3cm}

**Question 2**\newline
Write out the data frame that would be printed from this code.
```{r}
#| eval: false
planets |>
  select(name, mass, ring_system)
```


\newpage

**Question 3**\newline
Write out the data frame that would be printed from the following code.
```{r}
#| eval: false
planets |>
  select(name, mass, ring_system) |> 
  filter(ring_system == FALSE)
```

\vspace{4cm}

**Question 4**\newline
Describe how the following code behaves differently from the code in question 3.
```{r}
#| eval: false
planets <- planets |>
  select(name, mass, ring_system) |> 
  filter(ring_system == FALSE)
```

\vspace{3cm}

**Question 5**\newline
Which planets would be in the data frame printed by this code?
```{r}
#| eval: false
planets |>
  filter(length_of_day > 24 & surface_pressure <= 1)
```

\vspace{1.5cm}

**Question 6**\newline
Which planets would be in the data frame printed by this code?
```{r}
#| eval: false
planets |>
  filter(length_of_day > 24 | surface_pressure <= 1)
```

\vspace{1.5cm}

**Question 7**\newline
How many planets would be in the data frame printed by this code?
```{r}
#| eval: false
planets |>
  filter(mean_temp != "negative" | ring_system == FALSE)
```

\vspace{1.5cm}


**Question 8**\newline
Think about the code from question 7. Why did we have quotations around "negative" but not around `FALSE`?

\vspace{3cm}

**Question 9**\newline
Identify the issue with this code and correct it.

```{r}
#| eval: false
planets |>
  select(!name, !mean_temp)
```

\vspace{3cm}

**Question 10**\newline
Write out the data frame that would be printed from this code.

```{r}
#| eval: false
planets |>
  select(name, mass, ring_system) |> 
  filter(ring_system == FALSE) |> 
  mutate(mass_approx = round(mass))
```

\vspace{4cm}

**Question 11**\newline
What type of variable will `moons_per_mass` be (character, logical, or numeric)?
```{r}
#| eval: false
planets |> 
  mutate(moons_per_mass = n_moons / mass)
```

\vspace{1cm}

**Question 12**\newline
Identify the value of `mass_category` for each planet resulting from this code.
```{r}
#| eval: false
planets |> 
  mutate(mass_category = case_when(
    mass < 1 ~ "Small",
    mass < 10 ~ "Medium",
    TRUE ~ "Large"
  ))
```

\vspace{3cm}

**Question 13**\newline
I would like you to write one pipeline that does the following (in order):

1. Subsets the columns to only: `name`, `mass`, `n_moons`, `surface_pressure`.
2. Creates a new variable called `relative_day` which stores: "Short" if the length of a day is less than 24 hours, "Regular" if it exactly equals 24, and "Long" if it is longer than 24 hours.
3. Filters to only planets with "Short" days and more than 50 moons.