---
title: "Lab 2a"
subtitle: "Data basics and summary statistics"
lang: en
format: 
  pdf:
    echo: false
    message: false
    warning: false
    pdf-standard: ua-2
    geometry:
      - top=0.75in
      - left=1in
      - right=1in
    include-before-body:
      text: |
        \thispagestyle{fancy}
    include-in-header:
      text: |
        \usepackage{lastpage}
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{}
        \fancyhead[L]{DATA 200 Spring 2026}
        \fancyhead[R]{Hello Data Science Chapter 2}
        \fancyfoot[C]{\thepage\ of \pageref{LastPage}}
        \setlength{\headsep}{2pt}
        \RedeclareSectionCommand[afterskip=0.2em]{section}
        \RedeclareSectionCommand[afterskip=0.1em]{subsection}
        \RedeclareSectionCommand[afterskip=0.1em]{subsubsection}
---

\vspace*{-2.5cm}

```{r}
#| echo: false
library(tidyverse)
library(knitr)
options(scipen = 999)
```

**Team members:** 

Today's goal: Practice describing a dataset, distinguishing variable types, and choosing and computing useful summaries without using R.

## Question 1: parts of a dataset

```{r}
student_survey <- data.frame(
  student = c(101, 102, 103, 104, 105, 106, 107, 108),
  class = c("First year", "Junior", "Sophmore", "Senior", "First year", "Junior", "Senior", "Sophmore"),
  commute = c(10, 35, 0, 50, 15, 25, 5, 20),
  events_attended = c(1, 3, 0, 4, 2, 2, 5, 1),
  preference = c("Career panel", "Workshop", "Workshop", "Career panel", "Social", "Workshop", "Career panel", "Social"),
  plans_to_return = c("Yes", "No", "Yes", "No", "Yes", "No", "Yes", "Yes")
) |> 
  slice(1:5)
```


Suppose 5 students respond to a survey about campus events, and the data is recorded into an object named `student_survey`. Below is the same data presented in two formats: how the data can be presented to a human and how we could store the data on the computer using `R`.

```{r}
student_survey |> 
  kable(
    col.names = c("Student", "Class", "Commute (mins)", "Events attended", "Preferred event", "Plans to return")
  )
```

```{r}
#| echo: true
student_survey
```

### Question 1a

What is an observation in this dataset? Who or what does each observation describe?

\vspace{1cm}

### Question 1b

How many observations and how many variables are shown?

\vspace{1cm}


## Question 2

Classify each variable by its variable type and ideal `R` storage type. Be as specific as possible: numerical discrete, numerical continuous, categorical nominal, or categorical ordinal. 

| variable | statistical variable type | ideal `R` storage type |
|:---|:---------|:--------|
| `student` | | 
| | | |
| `class` | | |
| | | |
| `commute` | | |
| | | |
| `events_attended` | | |
| | | |
| `preferred` | | |
| | | |
| `plans_to_return` | | |
| | | |

\vspace*{-0.75cm}

## Question 3

Diego says, "`student` is numerical because it contains numbers." Do you agree? Explain what makes a variable numerical rather than merely stored with numbers.

\vspace{0.5in}

## Question 4: Summarizing a categorical variable

### Question 4a

Complete the frequency table for `preferred_event`.

| Preferred event | count | proportion | percent |
|:---|:------------------|:------------------|:------------------|
| Career panel | | | |
| | | | | 
| Social | | | |
| | | | | 
| Workshop | | | |
| | | | | 
| **Total** | | | |

\vspace*{-0.75cm}

### Question 4b

Write one complete sentence interpreting a percentage from your table in context.

\vspace{0.5in}

## Question 5: Summarizing a numerical variable

The commute times, already ordered, are: `r paste0(sort(student_survey$commute),  collapse = ", ")`.  

\vspace*{-0.5cm}

### Question 5a

Calculate each summary. Show your work as appropriate.

- Mean:

\vspace{0.25in}

- Median:

- Range (distance between minimum and maximum values):

- 40th percentile:

### Question 5b

Interpret the median commute time in context.

\vspace{0.5in}


### Question 5c

One student corrects their commute time from 50 minutes to 100 minutes. Without recalculating exact values, predict what happens to each statistic. Write **increases**, **decreases**, or **stays the same**.

- Mean:

- Median:

- Variance:

- Standard deviation:

- First Quartile

- IQR:

Which measure(s) are more resistant to this change of a single extreme value? Explain briefly.

\vspace{0.5in}

## Question 6

Match each research question to the most useful summary statistic. A statistic may be used more than once. Choices: count/proportion, mean, median, standard deviation, IQR.

- What is a typical number of events attended?

\vspace{0.5cm}

- Do most respondents plan to return to a campus event?

\vspace{0.5cm}


- How much do commute times typically vary around their mean?

\vspace{0.5cm}


- What is a typical commute time, when the data has one extremely large value? 


\vspace{0.5cm}

- How spread out is the middle 50% of commute times?

\vspace{0.5cm}



## Question 7

For each claim, decide whether the evidence supports it. If not, identify what is wrong.

- "The majority of surveyed students have attended a campus event."

\vspace{1in}

- "Because 60% of these respondents plan to return, 60% of all CSUCI students plan to return."

\vspace{1in}

- "The typical respondent attends 2 events."

\vspace{1in}

- "Students having a long commute makes them not plan to return to a campus event."

\vspace{1in}



