---
title: "Data 200: Week 7 EDA Instructions"
author: "Your name here"
format: 
  html:
    echo: true
    message: false
    warning: false
    toc: true
    embed-resources: true
---

```{r}
#| echo: false
library(janitor)
library(knitr)
library(tidyverse)
```

# Objective

For this investigation, you will explore how poverty and food-store access vary across counties in a US state of your choice. 
Each student must choose a different state, and California is reserved for the example. 

Use the California example to guide the appearance and organization of your results (I am not showing your my code).


# Wrangling the data

## Obtaining the data

1. Visit U.S. Department of Agriculture's (USDA's) [Food Environment Atlas ](https://ers.usda.gov/data-products/food-environment-atlas/data-access-and-documentation-downloads) and download the `Food Environment Atlas .csv Files` (current Version).

2. Move the 3 downloaded data files into your week 7 repo (two csv files and one txt file).

3. Open the `ReadMeFile2025.txt` and read it to learn about this data.


## Preparing your usda data

4. Notice I provided you with the code to read in and initially prepare your data. Remove `#| eval: false` from the provided chunk so it will run when you render your document. 

5. Run the provided code to import the data, clean the column names, and reshape the dataset. Examine both `usa_food_og` and `usa_food`; you will explain their differences in Reflection Q1.

6. Starting with usa_food, create a new object containing only the counties in your selected state. Give this object a descriptive name.

7. For your subsetted data, convert metro23 to a logical variable.

8. For your subsetted data, reorder the county factor by `pct_laccess_pop19`, the percentage of residents with low food-store access in 2019.

9. For your subsetted data, use `!ends_with("*")` with a `seelct()` to drop columns that are not for county level data.

## Preparing your map data

In order to create the map we have to first prepare the data to include latitude and longitude values for counties in your state.

10. Use the `map_data("county", region = )` function to load in a dataset of your state's geography, with region ==  the name of your state. Save this dataset into an object with a name of your choice.
11. We want to join this map data with our usda data by county name, but the county names are capitalized differently in each dataset. Convert the county names to title case and store them into a column named `county`.
12. Join the cleaned map data with the cleaned usda data, keeping all rows and columns. Save the joined dataset into a new object with a name of your choice.

## Analyzing your data

Use your food dataset with one row per county for these analyses. 
The joined map dataset contains repeated rows for each county’s coordinates and should be used only for mapping.

1. Group counties by metropolitan status and calculate each group’s median poverty rate in 2021 (povrate21) and median percentage of residents with low food-store access in 2019 (pct_laccess_pop19). Give the summary columns descriptive names. If values are missing, exclude them from the calculations.

2. Create a plot showing the relationship between poverty rate and low food-store access. Use county names as text labels to represent the observations using `geom_text()` instead of `geom_point()`. Include descriptive axis labels with units and years and specify alternative text.

3. Consult the variable documentation and choose one additional county-level variable that you think could be related to poverty. Do not choose a state-level measure repeated across counties. Explore the relationship between poverty and your chosen variable using an appropriate visualization, grouped summary statistics, or both.  Label your output clearly, including units and measurement years where applicable.

4. Select the county name and all variables you investigated: poverty rate, low food-store access, metropolitan status, and your additional variable. Pipe this into the `kable()` function to have a table display nicely in your html report.


## Creating the map

5. Use your joined map dataset to create a county-level map of your selected state. 

```{r}
#| eval: false
ggplot(
  ca_food_map_data, # replace this dataset with the name of yours
  aes(
    x = long, 
    y = lat, 
    group = county, 
    fill = povrate21
  )
) +
  geom_polygon(color = "white")
```


6. Change the colors to be from a color-blind friendly palette.

7. Add a descriptive plot title and legend title. 


## Reflection questions

Answer the reflection questions found at the end of our eda file.

Make sure to save your changes and push them to GitHub.
